---
title: "report"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Data preparation

We start by checking whether there are zero or near-zero variance variables in the dataset:

```{r}
library(caret)
library(randomForest)

dataset <- read.csv("pml-training.csv")
nzv <- nearZeroVar(dataset, saveMetrics = TRUE)
nzv
```

No variable has zero variance, but 60 have near-zero. To make computation faster, we try to work without near-zero variance variables:


```{r}
nzv <- nearZeroVar(dataset)
dataset <- dataset[, -nzv]
```

We remove columns that should not have predicting power: row numbers and time stamps (it's not a time series)

```{r}
drops <- c("X",
           "raw_timestamp_part_1",
           "raw_timestamp_part_2",
           "cvtd_timestamp")
dataset <- dataset[ , !names(dataset) %in% drops]
```

The randomForest algorithm we'll use later gives error for partially filled datasets: let's try to fit a random Forest model only using 'complete' variables:

```{r}
NAvars <- sapply(dataset, function(x) mean(is.na(x))) > 0.95
dataset <- dataset[NAvars==FALSE]
```

Finally, we split the training dataset into training and validation sets. We'll use the former to fit our model, and the later to validate it:

```{r}
set.seed(113)
inTrain = createDataPartition(y = dataset$classe, p = 0.7, list=FALSE)
training = dataset[inTrain,]
validation = dataset[-inTrain,]
```

## Random Forest

```{r}

set.seed(211)
ctrl <- trainControl(method = "cv",
                     verboseIter = FALSE)
mod_rf <- train(classe ~ .,
                method = "rf",
                data = training,
                trControl = ctrl,
                #centers and scales predictors
                preProc = c("center", "scale"))
```

We test our model on an out-of-sample validation test, different than the training one we fit our model on:

```{r}
randomForClasses <- predict(mod_rf, newdata = validation)
```

The random forest approach on the smaller but complete number of variables performs very well:

```{r}
confusionMatrix(randomForClasses, validation$classe)
```

Finally, here are the predictions for the provided testing set:

```{r}
assignment_testing <- read.csv("pml-testing.csv")
randomForClasses_test <- predict(mod_rf, newdata = assignment_testing)
randomForClasses_test
```